---
title: "Lesser Yellowlegs Integrated Population Model"
subtitle: "Anchorage breeding population — results summary"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 4.8
    fig-align: center
    tbl-cap-location: top
    css: report.css
execute:
  warning: false
  message: false
  echo: true
params:
  samples_path: null
---

```{r}
#| label: setup
#| include: false

library(coda)
library(gt)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| label: load
#| include: false

# Run from the project root regardless of where the .qmd lives.
proj <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
setwd(proj)

# Rebuild the data context (ipm_years, T, cjs_int_start, cjs, dp ...) exactly as
# the model saw it. SKIP_RUN keeps this to object definitions -- no refitting.
SKIP_RUN <- TRUE
source("scripts/LEYE_IPM.R")

# Locate the cached posterior.
#
# Use the CANONICAL filename, not a newest-match glob. A glob on
# LEYE_IPM_samples*.rds also matches companion and archive fits
# (vague-priors, time-varying-omega, dated caches), and picking whichever is
# newest silently builds the report from whichever model was last refitted.
# That fails loudly only if the parameter names happen to differ; if they
# match, the report renders cleanly from the wrong posterior.
CANONICAL <- "LEYE_IPM_samples.rds"
samples_path <- params$samples_path
if (is.null(samples_path)) {
  if (!file.exists(CANONICAL)) {
    stop(
      "No cached posterior at '", CANONICAL, "'. Fit the model first ",
      "(scripts/LEYE_IPM.R or scripts/LEYE_IPM_run_parallel.R), or pass ",
      "-P samples_path:<file>."
    )
  }
  samples_path <- CANONICAL
}

# The parallel wrapper's lapply() strips the mcmc.list class, and as.matrix()
# on the resulting plain list silently returns a 3x1 matrix instead of erroring.
# Re-wrap defensively so the report cannot be built on garbage.
samples <- readRDS(samples_path)

# Provenance guard. A cache fitted under a different model version renders
# perfectly well and simply shows the wrong numbers, so refuse rather than warn.
cache_status <- ipm_cache_status(samples)
if (cache_status == "stale") {
  stop(
    "'", samples_path, "' was fitted under a DIFFERENT version of the model ",
    "than the one in scripts/LEYE_IPM.R. Refit before rendering, or the report ",
    "will present the previous fit's results as current. (Comments and ",
    "formatting are excluded from the fingerprint, so this reflects a real ",
    "change to structure, constants, priors or data.)"
  )
}

# as.mcmc.list() drops attributes, so read provenance before converting. The
# parallel wrapper's lapply() used to strip the mcmc.list class, and as.matrix()
# on a plain list silently returns a 3 x 1 matrix -- re-wrap defensively.
if (!inherits(samples, "mcmc.list")) samples <- coda::as.mcmc.list(samples)
mat <- as.matrix(samples)

n_chains <- coda::nchain(samples)
n_draws <- nrow(mat)
stopifnot(n_draws > 100) # guard against the silent-truncation failure above

# Priors on recruitment, as declared in scripts/LEYE_IPM.R. Kept here so the
# prior-vs-posterior comparison below cannot drift from the fitted model.
PRIOR_PI <- c(a = 2.268, b = 13.610)
PRIOR_KAPPA <- c(a = 6.054, b = 6.309)

# omega is a scalar in some fits and a year-indexed vector in others. Resolve
# once here so the rest of the report does not care which it is.
omega_tv <- any(grepl("^omega\\[", colnames(mat)))
omega_draws <- if (omega_tv) mat[, "omega_bar"] else mat[, "omega"]

# tLTRE. Computed here from the SAME posterior loaded above rather than read
# from LEYE_tLTRE_results.rds, so the decomposition cannot silently belong to a
# different fit than the numbers everywhere else in this report. Sourcing with
# SKIP_TLTRE_RUN loads the functions without running the script's own report.
SKIP_TLTRE_RUN <- TRUE
source("scripts/LEYE_tLTRE.R")
tl <- tltre_series(mat, T)
tltre_check(tl) # stops the render if the exact identity ever breaks
tl_level <- tltre_level(tl)
tl_var <- tltre_variance(tl)

driver_labs <- c(
  phi = "Adult survival (φ)",
  omega = "Immigration (ω)",
  f_t = "Productivity, year t",
  f_lag = "Productivity, year t−1",
  rho = "Population structure (ρ)",
  psi1 = "Recruitment at age 1 (ψ1)",
  psi2 = "Recruitment at age 2 (ψ2)"
)
# NB: plain "ψ1" not "ψ₁" -- the Unicode subscript has no glyph in the default
# plotting font and renders as a tofu box on the axis.

q <- function(x) {
  c(
    mean = mean(x),
    med = median(x),
    lwr = quantile(x, .025, names = FALSE),
    upr = quantile(x, .975, names = FALSE)
  )
}
qq <- function(stem, idx) {
  t(vapply(paste0(stem, "[", idx, "]"), function(cn) q(mat[, cn]), numeric(4)))
}

# House table style: one call so every table in the report matches.
tbl <- function(x, title = NULL, subtitle = NULL, dec = 3) {
  g <- gt(x)
  if (!is.null(title)) g <- tab_header(g, title = title, subtitle = subtitle)
  g |>
    fmt_number(columns = where(is.numeric), decimals = dec) |>
    opt_row_striping() |>
    tab_options(
      table.font.size = px(13),
      heading.title.font.size = px(15),
      heading.subtitle.font.size = px(12),
      column_labels.font.weight = "600",
      data_row.padding = px(4),
      table.border.top.style = "none"
    )
}

# Shared plot look.
theme_leye <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey35", size = 11),
      plot.caption = element_text(colour = "grey45", hjust = 0, size = 9),
      axis.title = element_text(colour = "grey25")
    )
}
COL_AD <- "#1f4e79"
COL_REC <- "#2e8b57"
COL_IMM <- "#c0504d"

trans_yr <- ipm_years[-T]
trans_lab <- paste0(trans_yr, "–", trans_yr + 1)
```

::: {.callout-note appearance="simple"}
## How to read this report

Abundance is estimated on a **relative** scale — the counts are a volunteer index,
not a census. Read the *shape* of the trajectory and the growth rate $\lambda$;
the absolute number of birds is not identified and is not reported here.

Recruitment (`psi1`, `psi2`) is a **composite** probability: survive **and**
return to breed locally. It is not juvenile survival, and should not be compared
against published juvenile survival estimates.
:::

# Overview

This integrated population model combines three data sources into one likelihood
so that each informs a shared set of demographic rates.

```{r}
#| label: tbl-sources
#| echo: false

data.frame(
  Source = c("Abundance", "Survival", "Productivity"),
  Data = c(
    "Volunteer wetland counts",
    "Adult mark–resight (CJS)",
    "Nest productivity"
  ),
  Years = c("2014–2025", "2010–2026", "2018–2026"),
  Detail = c(
    sprintf("%d records, %d wetlands, 3 weekly visits", nCount, nWet),
    sprintf("%d individuals, sex + tag-type covariates", cjs$nind),
    sprintf("%d nests, chicks per nest", nProd)
  )
) |>
  tbl("Data sources") |>
  cols_align("left")
```

Recruitment carries a **two-year lag**: chicks are absent from the breeding site
until they recruit, and the 1990s chick data show recruitment split roughly evenly
between ages 1 and 2. Recruits enter the breeding class directly, so there is no
separate juvenile stage — volunteer counts cannot distinguish non-breeding age-1
birds from breeding adults.

$$
R_t = N_{t-1}\,\frac{f_{t-1}}{2}\,\psi_1 \;+\; N_{t-2}\,\frac{f_{t-2}}{2}\,\psi_2
$$

# Headline results

```{r}
#| label: headline
#| echo: false

lam <- rowMeans(mat[, paste0("lambda[", 1:(T - 1), "]")])
geolam <- exp(mean(log(lam)))
geolam_ci <- exp(quantile(log(lam), c(.025, .975)))
Rm <- rowSums(mat[, paste0("N_rec[", 3:T, "]")])
Im <- rowSums(mat[, paste0("Imm[", 3:T, "]")])
share <- Rm / (Rm + Im)

card <- function(value, label, sub, colour) {
  sprintf(
    '<div class="kpi" style="border-top-color:%s"><div class="kpi-v">%s</div>
     <div class="kpi-l">%s</div><div class="kpi-s">%s</div></div>',
    colour, value, label, sub
  )
}
htmltools::HTML(paste0(
  '<div class="kpi-row">',
  card(
    sprintf("%.3f", geolam), "Population growth (λ)",
    sprintf("%.3f – %.3f &nbsp;|&nbsp; %.1f%%/yr", geolam_ci[1], geolam_ci[2], 100 * (geolam - 1)),
    "#c0504d"
  ),
  card(
    sprintf("%.2f", mean(mat[, "phi_ad[1]"])), "Adult survival",
    "annual, untagged female baseline", COL_AD
  ),
  card(
    sprintf("%.3f", mean(mat[, "pi_rec"])), "Local recruitment",
    sprintf("ψ₁ %.3f &nbsp; ψ₂ %.3f", mean(mat[, "psi1"]), mean(mat[, "psi2"])),
    COL_REC
  ),
  card(
    sprintf("%.3f", mean(omega_draws)), "Immigration rate (ω)",
    sprintf("%.0f%% of annual gains", 100 * (1 - mean(share))),
    COL_IMM
  ),
  "</div>"
))
```

The population is declining at roughly
`r sprintf("%.1f%%", abs(100 * (geolam - 1)))` per year. Adult survival is the
best-determined rate in the model. Local recruitment is low, and a substantial
share of annual gains is attributable to immigration — but see
[Recruitment versus immigration] for why that split remains uncertain.

# Population trajectory

```{r}
#| label: fig-trajectory
#| fig-cap: "Estimated breeding population on a relative scale. Ribbon is the 95% credible interval. The absolute level is not identified — read the shape, not the height."

Nt <- as.data.frame(qq("N_tot", 1:T))
Nt$year <- ipm_years

ggplot(Nt, aes(year, med)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = COL_AD, alpha = .15) +
  geom_line(colour = COL_AD, linewidth = .9) +
  geom_point(colour = COL_AD, size = 2.2) +
  scale_x_continuous(breaks = ipm_years) +
  labs(
    title = "Estimated population trajectory",
    subtitle = "Relative scale — interpret shape and growth rate, not absolute abundance",
    x = NULL, y = "Latent abundance (relative)"
  ) +
  theme_leye()
```

```{r}
#| label: fig-lambda
#| fig-cap: "Annual population growth rate. Values below 1 indicate decline. The dashed line marks a stable population."

lamd <- as.data.frame(qq("lambda", 1:(T - 1)))
lamd$transition <- factor(trans_lab, levels = trans_lab)

ggplot(lamd, aes(transition, med)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey45") +
  geom_linerange(aes(ymin = lwr, ymax = upr), colour = COL_AD, linewidth = .7) +
  geom_point(colour = COL_AD, size = 2.2) +
  labs(
    title = "Population growth rate by transition",
    subtitle = sprintf("Geometric mean λ = %.3f (%.3f–%.3f)", geolam, geolam_ci[1], geolam_ci[2]),
    x = NULL, y = expression(lambda)
  ) +
  theme_leye() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Adult survival

`phi_ad` is the baseline rate for **untagged females** — the reference level, and
the rate that drives the population process model, since the population is
overwhelmingly untagged.

```{r}
#| label: fig-survival
#| fig-cap: "Baseline annual adult survival (untagged female). Open points mark transitions with no direct mark–resight data, informed only through the process model."

ad <- as.data.frame(qq("phi_ad", 1:(T - 1)))
ad$transition <- factor(trans_lab, levels = trans_lab)
ad$has_cjs <- trans_yr %in% cjs_int_start

ggplot(ad, aes(transition, med)) +
  geom_linerange(aes(ymin = lwr, ymax = upr), colour = COL_AD, linewidth = .7) +
  geom_point(aes(shape = has_cjs), colour = COL_AD, fill = "white", size = 2.4) +
  scale_shape_manual(values = c(`TRUE` = 21, `FALSE` = 1), guide = "none") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Baseline adult survival",
    subtitle = "Untagged, female reference level",
    x = NULL, y = "Annual apparent survival"
  ) +
  theme_leye() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Covariate effects

Effects are shown as odds ratios against the reference levels (female, untagged).
An odds ratio below 1 means lower survival — or lower resighting probability —
than the reference.

```{r}
#| label: fig-forest
#| fig-cap: "Sex and tag-type effects on survival and resighting. Bars are 95% credible intervals; the dashed line at 1 is no effect."
#| fig-height: 5.2

orq <- function(col, label, panel) {
  x <- exp(mat[, col])
  data.frame(
    term = label, panel = panel,
    med = median(x),
    lwr = quantile(x, .025, names = FALSE),
    upr = quantile(x, .975, names = FALSE),
    pneg = mean(mat[, col] < 0)
  )
}
fo <- rbind(
  orq("bSex_phi[2]", "Male (vs female)", "Adult survival"),
  orq("bSex_phi[3]", "Unknown sex (vs female)", "Adult survival"),
  orq("bTag_phi[2]", "Geolocator (vs untagged)", "Adult survival"),
  orq("bTag_phi[3]", "GPS (vs untagged)", "Adult survival"),
  orq("bTag_p[2]", "Geolocator (vs untagged)", "Resighting"),
  orq("bTag_p[3]", "GPS (vs untagged)", "Resighting")
)
fo$term <- factor(fo$term, levels = rev(unique(fo$term)))

ggplot(fo, aes(med, term)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey45") +
  geom_linerange(aes(xmin = lwr, xmax = upr), colour = COL_AD, linewidth = .7) +
  geom_point(colour = COL_AD, size = 2.4) +
  geom_text(aes(x = upr, label = sprintf("P(<1)=%.2f", pneg)),
    hjust = -0.15, size = 2.9, colour = "grey40"
  ) +
  scale_x_log10(expand = expansion(mult = c(.05, .28))) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(
    title = "Covariate effects (odds-ratio scale)",
    subtitle = "Reference levels: female, untagged",
    x = "Odds ratio (log scale)", y = NULL
  ) +
  theme_leye() +
  theme(panel.grid.major.x = element_line(colour = "grey92"))
```

```{r}
#| label: tbl-groups
#| echo: false

base <- qlogis(mat[, "phi_ad[1]"])
bS <- cbind(0, mat[, "bSex_phi[2]"], mat[, "bSex_phi[3]"])
bT <- cbind(0, mat[, "bTag_phi[2]"], mat[, "bTag_phi[3]"])
grid <- expand.grid(tag = c("None", "Geolocator", "GPS"), sex = c("Female", "Male", "Unknown"))
grid$estimate <- NA_real_
grid$lwr <- NA_real_
grid$upr <- NA_real_
for (i in seq_len(nrow(grid))) {
  v <- plogis(base + bS[, match(grid$sex[i], c("Female", "Male", "Unknown"))] +
    bT[, match(grid$tag[i], c("None", "Geolocator", "GPS"))])
  grid[i, c("estimate", "lwr", "upr")] <- c(mean(v), quantile(v, c(.025, .975)))
}
grid |>
  select(Sex = sex, Tag = tag, Estimate = estimate, Lower = lwr, Upper = upr) |>
  arrange(Sex, Tag) |>
  tbl(
    "Adult survival by sex and tag type",
    "First transition. GPS × unknown-sex has no supporting data."
  ) |>
  cols_align("left", columns = c(Sex, Tag))
```

# Recruitment and immigration

## Local recruitment

`psi1` and `psi2` are the probabilities that a fledged chick survives **and**
returns to breed locally, at age 1 and age 2 respectively. Their priors come from
a separate analysis of 1990s chick cohorts
(`scripts/LEYE_juv90_module.R`) and are informative but free to update.

```{r}
#| label: tbl-recruit
#| echo: false

data.frame(
  Parameter = c("pi_rec (total)", "kappa (fraction at age 1)", "psi1 (age 1)", "psi2 (age 2)"),
  Mean = c(mean(mat[, "pi_rec"]), mean(mat[, "kappa"]), mean(mat[, "psi1"]), mean(mat[, "psi2"])),
  Lower = c(
    quantile(mat[, "pi_rec"], .025), quantile(mat[, "kappa"], .025),
    quantile(mat[, "psi1"], .025), quantile(mat[, "psi2"], .025)
  ),
  Upper = c(
    quantile(mat[, "pi_rec"], .975), quantile(mat[, "kappa"], .975),
    quantile(mat[, "psi1"], .975), quantile(mat[, "psi2"], .975)
  )
) |>
  tbl("Local recruitment", "Composite survival × philopatry — not annual survival") |>
  cols_align("left", columns = Parameter)
```

```{r}
#| label: fig-prior-post
#| fig-cap: "Prior versus posterior for the two recruitment parameters. Movement away from the prior shows the count data carrying information; overlap shows the estimate resting on the 1990s analysis."
#| fig-height: 3.6

grid_x <- seq(0, 1, length.out = 512)
dens <- function(nm, a, b, lab) {
  po <- density(mat[, nm], from = 0, to = 1)
  rbind(
    data.frame(x = grid_x, y = dbeta(grid_x, a, b), which = "Prior", par = lab),
    data.frame(x = po$x, y = po$y, which = "Posterior", par = lab)
  )
}
dd <- rbind(
  dens("pi_rec", PRIOR_PI["a"], PRIOR_PI["b"], "pi_rec (total recruitment)"),
  dens("kappa", PRIOR_KAPPA["a"], PRIOR_KAPPA["b"], "kappa (fraction at age 1)")
)

ggplot(dd, aes(x, y, colour = which, fill = which)) +
  geom_area(position = "identity", alpha = .18, colour = NA) +
  geom_line(linewidth = .8) +
  facet_wrap(~par, scales = "free") +
  scale_colour_manual(values = c(Prior = "grey55", Posterior = COL_REC), name = NULL) +
  scale_fill_manual(values = c(Prior = "grey55", Posterior = COL_REC), name = NULL) +
  labs(title = "Prior versus posterior", x = NULL, y = "Density") +
  theme_leye() +
  theme(legend.position = "top", axis.text.y = element_blank())
```

::: {.callout-important appearance="simple"}
The counts inform the **level** of recruitment but not its **timing**. `pi_rec`
moves appreciably from its prior; `kappa` barely moves. The age-1 versus age-2
split therefore rests entirely on the 1990s chick histories, not on the count
data — and those cohorts are partly simulated (see [Caveats]).
:::

## Recruitment versus immigration {#recruitment-versus-immigration}

```{r}
#| label: fig-gains
#| fig-cap: "Annual gains to the breeding population, split into locally recruited birds and immigrants. Bar heights are posterior means."

gains <- rbind(
  data.frame(
    year = ipm_years[3:T], source = "Local recruits",
    value = colMeans(mat[, paste0("N_rec[", 3:T, "]")])
  ),
  data.frame(
    year = ipm_years[3:T], source = "Immigrants",
    value = colMeans(mat[, paste0("Imm[", 3:T, "]")])
  )
)

ggplot(gains, aes(year, value, fill = source)) +
  geom_col(position = "stack", width = .7) +
  scale_fill_manual(values = c("Local recruits" = COL_REC, "Immigrants" = COL_IMM), name = NULL) +
  scale_x_continuous(breaks = ipm_years[3:T]) +
  labs(
    title = "Sources of annual gains",
    subtitle = sprintf(
      "Local recruitment accounts for %.0f%% of gains (95%% CrI %.0f–%.0f%%)",
      100 * mean(share), 100 * quantile(share, .025), 100 * quantile(share, .975)
    ),
    x = NULL, y = "Birds added"
  ) +
  theme_leye()
```

::: {.callout-warning appearance="simple"}
**This split is uncertain.** Recruitment and immigration both add birds,
and a relative-scale count index cannot fully separate them. The credible interval
on the recruitment share spans
`r sprintf("%.0f%%", 100 * quantile(share, .025))` to
`r sprintf("%.0f%%", 100 * quantile(share, .975))`. Informative priors on
recruitment narrow this but do not resolve it. Treat the point estimate as
indicative, not as a finding.
:::

# Productivity

Chicks per nest are modelled in **two parts**: whether the nest succeeded at all,
and how many of a four-egg clutch fledged given that it did.

```{r}
#| label: prod-params
#| echo: false
#| results: asis

s_bar <- rowMeans(mat[, paste0("s_nest[", 1:(T + 1), "]")])
cat(sprintf(
  paste0(
    "Nest success averages **%.2f** (95%% CrI %.2f–%.2f) and, given success, ",
    "each egg fledges with probability **%.2f** (%.2f–%.2f). Mean chicks per ",
    "nest is the product of the two with clutch size, so `f_rate` keeps the ",
    "same meaning it had under the earlier single-distribution model and the ",
    "recruitment equation is unchanged.\n\n"
  ),
  mean(s_bar), quantile(s_bar, .025), quantile(s_bar, .975),
  mean(mat[, "q_brood"]), quantile(mat[, "q_brood"], .025),
  quantile(mat[, "q_brood"], .975)
))
```

::: {.callout-note appearance="simple"}
This was initially modeled using a single Poisson draw for chicks per nest, but failed the posterior
predictive check (see @sec-gof). The observed counts are bimodal — 0, 2, 3 or 4 chicks, with
**no nest producing exactly one** — because a four-egg-clutch shorebird usually either
loses the nest entirely or fledges a near-full brood. A Poisson cannot generate
that shape at any mean; at the observed mean it expects roughly 16 single-chick
nests. The two-part model reproduces the observed frequencies closely.

This also tightened productivity in years with few monitored nests, where the
old model fitted a free annual mean to as few as two or three nests. The two-part
version borrows strength through the shared clutch size and fledging rate.
:::

```{r}
#| label: fig-fec
#| fig-cap: "Estimated chicks per pair. The final year is estimated from productivity data that has no matching count year and does not feed the population process."

fr <- as.data.frame(qq("f_rate", 1:(T + 1)))
fr$year <- c(ipm_years, max(ipm_years) + 1)
fr$in_model <- c(rep(TRUE, T), FALSE)

ggplot(fr, aes(year, med)) +
  geom_linerange(aes(ymin = lwr, ymax = upr), colour = "grey55", linewidth = .7) +
  geom_point(aes(colour = in_model), size = 2.2) +
  scale_colour_manual(
    values = c(`TRUE` = COL_AD, `FALSE` = "grey60"),
    labels = c(`TRUE` = "In process model", `FALSE` = "Beyond count series"), name = NULL
  ) +
  scale_x_continuous(breaks = fr$year) +
  labs(title = "Productivity", x = NULL, y = "Chicks per pair") +
  theme_leye() +
  theme(legend.position = "top")
```

# What drives population change

A transient LTRE (tLTRE), following Koons et al. (2016, 2017). Two questions that
sound similar and have different answers:

- **Level** — what *sustains* the population? Mean λ splits across survival,
  recruitment and immigration.
- **Variance** — what *drives year-to-year variation* in λ?

Everything below is computed per posterior draw, so each contribution carries a
credible interval rather than being a point estimate at posterior means.

## What sustains the population

```{r}
#| label: fig-tltre-level
#| fig-cap: "Additive contributions to mean λ. Bars are 95% credible intervals; percentages are each pathway's share of λ."
#| fig-height: 2.9

lev <- as.data.frame(tl_level)
lev$pathway <- factor(rownames(tl_level), levels = rev(rownames(tl_level)))
lam_mean <- attr(tl_level, "lambda")[["mean"]]

ggplot(lev, aes(contribution, pathway)) +
  geom_linerange(aes(xmin = lwr, xmax = upr), colour = COL_AD, linewidth = .7) +
  geom_point(colour = COL_AD, size = 2.6) +
  # label at the interval's right edge, not the point, or it lands on the bar
  geom_text(aes(x = upr, label = sprintf("%.1f%%", pct_of_lambda)),
    hjust = -0.25, size = 3.1, colour = "grey35"
  ) +
  scale_x_continuous(expand = expansion(mult = c(.05, .18))) +
  labs(
    title = "Contributions to mean λ",
    subtitle = sprintf("Percentages are share of λ = %.3f", lam_mean),
    x = "Contribution to λ", y = NULL
  ) +
  theme_leye()
```

Survival is the dominant pathway. Recruitment and immigration contribute
comparable, smaller amounts — but see the caveat on separating them below.

## What drives year-to-year variation

::: {.callout-important appearance="simple"}
**Most variation in λ is demographic stochasticity, not environmental variation
in vital rates.** With roughly 25 birds, Poisson noise in who happens to survive,
recruit or arrive is a first-order term. A conventional LTRE reports only the
driver contributions and would present a fraction of the variance as though it
were the whole story.
:::

```{r}
#| label: tbl-tltre-partition
#| echo: false

pp <- as.data.frame(tl_var$partition_pct)
pp$Component <- c(
  "Driver-explained", "Demographic stochasticity", "Cross term"
)
pp[c("Component", "mean", "lwr", "upr")] |>
  setNames(c("Component", "Share of Var(λ) %", "Lower", "Upper")) |>
  tbl(
    "Partition of Var(λ)",
    "The three sum to 100%. Individual shares can exceed it because the cross term is negative.",
    dec = 1
  ) |>
  cols_align("left", columns = Component)
```

The cross term is non-zero because population structure (`ρ`) depends on realized
abundance and so carries the same Poisson noise — the smooth and stochastic parts
of λ are not independent.

```{r}
#| label: fig-tltre-drivers
#| fig-cap: "tLTRE contributions to the driver-explained portion of Var(λ). Bars are 95% credible intervals. Parameters with no year index contribute exactly zero by construction."
#| fig-height: 3.4

ct <- as.data.frame(tl_var$contributions)
ct$driver <- rownames(tl_var$contributions)
ct$label <- driver_labs[ct$driver]
ct$structural_zero <- ct$mean == 0 & ct$lwr == 0 & ct$upr == 0
ct$label <- factor(ct$label, levels = rev(ct$label))

ggplot(ct, aes(mean, label)) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_linerange(aes(xmin = lwr, xmax = upr, colour = structural_zero), linewidth = .7) +
  geom_point(aes(colour = structural_zero), size = 2.6) +
  scale_colour_manual(
    values = c(`FALSE` = COL_AD, `TRUE` = "grey65"), guide = "none"
  ) +
  # phi's upper bound sits past the default break, so widen or it gets clipped
  scale_x_continuous(expand = expansion(mult = c(.04, .08))) +
  labs(
    title = "Contributions to driver-explained variance",
    subtitle = sprintf(
      "Drivers explain %.0f%% of Var(λ); the rest is demographic noise",
      tl_var$partition_pct["driver_explained", "mean"]
    ),
    x = "Contribution to Var(λ)", y = NULL
  ) +
  theme_leye() +
  theme(panel.grid.major.x = element_line(colour = "grey93"))
```

```{r}
#| label: tltre-zeros
#| echo: false
#| results: asis

zeros <- rownames(tl_var$contributions)[
  tl_var$contributions[, "mean"] == 0 & tl_var$contributions[, "upr"] == 0
]
if (length(zeros)) {
  cat(
    "::: {.callout-note appearance='simple'}\n",
    "**", paste(driver_labs[zeros], collapse = ", "), "** contribute exactly ",
    "zero — they are single parameters with no year index, so they cannot vary ",
    "between years *by construction, not by evidence*. A time-varying ",
    "immigration rate was fitted and rejected: it absorbed Poisson noise rather ",
    "than detecting signal (see the README). Separating recruitment from ",
    "immigration needs data that distinguishes chick-origin returns from newly ",
    "arrived unbanded adults.\n:::\n",
    sep = ""
  )
}
```

# Model checking

## Convergence

```{r}
#| label: tbl-rhat
#| echo: false

keyp <- intersect(
  c(
    "pi_rec", "kappa", "psi1", "psi2", "omega", "omega_bar", "sd_omega",
    "phi_pre", "phi_post",
    "mu_s", "q_brood", "sd_s", "lp_base", "logB0",
    "bSex_phi[2]", "bSex_phi[3]", "bTag_phi[2]", "bTag_phi[3]",
    "bTag_p[2]", "bTag_p[3]"
  ),
  colnames(mat)
)
gd <- gelman.diag(samples[, keyp], multivariate = FALSE)$psrf
rh <- data.frame(Parameter = rownames(gd), Rhat = gd[, 1], Upper = gd[, 2])

tbl(rh, "Convergence diagnostics", sprintf("%d chains, %s pooled draws", n_chains, format(n_draws, big.mark = ","))) |>
  cols_align("left", columns = Parameter) |>
  data_color(
    columns = Rhat,
    fn = function(x) ifelse(x > 1.05, "#f8d7da", "#d4edda")
  ) |>
  tab_source_note("Values below 1.05 indicate adequate mixing.")
```

```{r}
#| label: logb0-note
#| echo: false
#| results: asis

if (any(gd[rownames(gd) == "logB0", 1] > 1.05)) {
  cat(
    "::: {.callout-note appearance='simple'}\n",
    "`logB0` is expected to converge poorly. It is the count-model intercept, ",
    "which is non-identifiable against the absolute level of the population — ",
    "only their sum enters the linear predictor. This is structural and does ",
    "not affect $\\lambda$ or any rate reported above.\n:::\n",
    sep = ""
  )
}
```

## Goodness of fit {#sec-gof}

An IPM can fit one data source well and another badly, and a single pooled
statistic would hide that. Each stream is therefore checked separately, using a
Freeman–Tukey discrepancy between the observed data and replicates drawn from
the fitted model. The Bayesian *p*-value is P(replicated > observed); values near
0.5 indicate the model reproduces that stream, values near 0 or 1 indicate misfit.

```{r}
#| label: tbl-gof
#| echo: false

gof <- data.frame(
  Stream = c("Counts", "Productivity", "Mark–resight"),
  p = c(
    mean(mat[, "FT_count_rep"] > mat[, "FT_count_obs"]),
    mean(mat[, "FT_prod_rep"] > mat[, "FT_prod_obs"]),
    mean(mat[, "FT_cjs_rep"] > mat[, "FT_cjs_obs"])
  )
)
names(gof)[2] <- "Bayesian p"
tbl(gof, "Posterior predictive checks", dec = 3) |>
  cols_align("left", columns = Stream) |>
  data_color(
    columns = `Bayesian p`,
    fn = function(x) ifelse(x < 0.05 | x > 0.95, "#f8d7da", "#d4edda")
  )
```

::: {.callout-note appearance="simple"}
This check earned its keep. The productivity component originally used a single
Poisson and returned *p* = 0.000 — decisive misfit — which led to the two-part
nest-success × brood-size model now in use.

Worth noting for anyone repeating this: the productivity data have
**variance/mean = 1.18**, so a routine overdispersion check would have passed
them. The misfit was in *shape*, not variance, and only a discrepancy measure
sensitive to the whole distribution detected it.
:::

## Prior sensitivity {#sec-prior-sens}

Recruitment carries informative priors derived from the 1990s chick analysis, so
the obvious question is whether the result is data-driven or prior-driven. The
comparison below refits the model with flat `Beta(1, 1)` priors on both
recruitment parameters, changing nothing else.

```{r}
#| label: tbl-prior-sens
#| echo: false
#| results: asis

sens_path <- "LEYE_IPM_samples_vaguepriors.rds"
if (file.exists(sens_path)) {
  sv <- readRDS(sens_path)
  # Checked on the STRUCTURAL fingerprint: this fit is meant to differ in the
  # recruitment priors, but not in anything else.
  sens_status <- ipm_cache_status(sv, key = "structural")
  sens_priors <- attr(sv, "ipm_provenance")$priors
  if (!inherits(sv, "mcmc.list")) sv <- coda::as.mcmc.list(sv)
  mv <- as.matrix(sv)
  fmt <- function(m, p) {
    sprintf("%.3f (%.3f–%.3f)", mean(m[, p]), quantile(m[, p], .025), quantile(m[, p], .975))
  }
  gl <- function(m) {
    l <- rowMeans(m[, paste0("lambda[", 1:(T - 1), "]")])
    sprintf("%.3f (%.3f–%.3f)", exp(mean(log(l))), exp(quantile(log(l), .025)), exp(quantile(log(l), .975)))
  }
  ps <- data.frame(
    Parameter = c("pi_rec (recruitment level)", "kappa (fraction at age 1)", "λ (geometric mean)"),
    Informative = c(fmt(mat, "pi_rec"), fmt(mat, "kappa"), gl(mat)),
    Vague = c(fmt(mv, "pi_rec"), fmt(mv, "kappa"), gl(mv))
  )
  print(tbl(ps, "Informative vs vague recruitment priors") |>
    cols_align("left", columns = everything()))

  if (sens_status == "stale") {
    cat(
      "\n\n::: {.callout-warning appearance='simple'}\n",
      "**This comparison is out of date.** The companion fit was produced ",
      "under a different version of the model than the one generating the rest ",
      "of this report, so the two columns are not directly comparable. Refit it ",
      "with the recruitment hyperparameters set to 1 and re-render.\n:::\n",
      sep = ""
    )
  } else if (sens_status == "unstamped") {
    cat(
      "\n\n::: {.callout-note appearance='simple'}\n",
      "The companion fit predates the provenance stamp, so it cannot be ",
      "verified as matching the current model. Treat the comparison as ",
      "indicative and refit when convenient.\n:::\n",
      sep = ""
    )
  } else if (!is.null(sens_priors)) {
    cat(sprintf(
      "\n\nCompanion fitted with Beta(%g, %g) on `pi_rec` and Beta(%g, %g) on `kappa`.\n",
      sens_priors[["a_pi"]], sens_priors[["b_pi"]],
      sens_priors[["a_kap"]], sens_priors[["b_kap"]]
    ))
  }
} else {
  cat(
    "*Companion fit `", sens_path, "` not found. To regenerate: refit with ",
    "`a_pi = b_pi = a_kap = b_kap = 1` in `constants` — the prior ",
    "hyperparameters are passed as constants precisely so this is a config ",
    "change rather than a second copy of the model.*\n",
    sep = ""
  )
}
```

::: {.callout-important appearance="simple"}
**The recruitment level is not an artifact of the prior; the timing is.**

Under a flat prior `pi_rec` barely moves, so the count data pull it to the same
place either way. `kappa` also keeps its central value but its interval widens
dramatically — the age-1 versus age-2 split rests entirely on the 1990s chick
histories, not on anything in the modern data. Since those cohorts are partly
simulated (see [Caveats]), the recruitment *schedule* is the weakest link in
this analysis, while the recruitment *rate* is on firmer ground.
:::

# Caveats {#caveats}

::: {.callout-important}
## Read before citing any number above

**The 1990s chick cohorts are partly simulated.** Recruitment priors derive from
`scripts/LEYE_juv90_module.R`, whose cohort denominators are real for 1996 only.
For 1995, 1997 and 1998 the number of broods is a random draw with an explicitly
arbitrary mean. Recruitment probability scales inversely with those denominators.

**Recruitment and immigration are confounded**, as described above.

**Abundance is relative.** Do not report absolute population size.

**`psi` is not survival.** It is survival combined with returning to breed
locally. Birds that survive but settle elsewhere are excluded; immigration is
modelled separately through `omega`.

**Counts end in 2025** while survival and productivity run to 2026. The extra
survival interval and productivity year are estimated in dedicated pooled
parameters rather than being discarded, and do not drive the trajectory.
:::

# Reproducibility

```{r}
#| label: repro
#| echo: false

prov <- attr(readRDS(samples_path), "ipm_provenance")
data.frame(
  Item = c(
    "Posterior file", "Chains", "Pooled draws", "Model script",
    "Fitted", "Model fingerprint", "Report rendered"
  ),
  Value = c(
    basename(samples_path),
    as.character(n_chains),
    format(n_draws, big.mark = ","),
    "scripts/LEYE_IPM.R",
    if (is.null(prov)) "not recorded" else format(prov$fitted, "%Y-%m-%d %H:%M"),
    if (is.null(prov)) "not recorded" else substr(prov$full, 1, 12),
    format(Sys.time(), "%Y-%m-%d %H:%M")
  )
) |>
  tbl("Run details") |>
  cols_align("left")
```

```{r}
#| label: cache-provenance
#| echo: false
#| results: asis

if (cache_status == "unstamped") {
  cat(
    "::: {.callout-note appearance='simple'}\n",
    "**This posterior predates the provenance stamp.** Caches are now ",
    "fingerprinted against the model that produced them, and rendering aborts ",
    "if they disagree — but that check cannot be applied retrospectively, so ",
    "this file's provenance is asserted rather than verified. The next refit ",
    "stamps it and the guard becomes active.\n:::\n",
    sep = ""
  )
} else {
  cat(
    "::: {.callout-note appearance='simple'}\n",
    "Posterior verified against the current model: structure, constants, ",
    "priors and data all match. Comments and formatting are excluded from the ",
    "fingerprint, so re-documenting the model does not invalidate a fit.\n:::\n",
    sep = ""
  )
}
```

To reproduce: fit the model with `scripts/LEYE_IPM_run_parallel.R` (or
`scripts/LEYE_IPM.R` for a serial run), then render this report with

```bash
quarto render reports/leye_ipm_report.qmd
```

<details>
<summary>Session information</summary>

```{r}
#| label: sessioninfo
#| echo: false
sessionInfo()
```

</details>
