---
title: "Lesser Yellowlegs Integrated Population Model"
subtitle: "Anchorage breeding population — results summary"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 4.8
    fig-align: center
    tbl-cap-location: top
    css: report.css
execute:
  warning: false
  message: false
  echo: true
params:
  samples_path: null
---

```{r}
#| label: setup
#| include: false

library(coda)
library(gt)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| label: load
#| include: false

# Run from the project root regardless of where the .qmd lives.
proj <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
setwd(proj)

# Rebuild the data context (ipm_years, T, cjs_int_start, cjs, dp ...) exactly as
# the model saw it. SKIP_RUN keeps this to object definitions -- no refitting.
SKIP_RUN <- TRUE
source("scripts/LEYE_IPM.R")

# Locate the cached posterior. Default: most recent LEYE_IPM_samples*.rds.
samples_path <- params$samples_path
if (is.null(samples_path)) {
  cand <- list.files(".", "^LEYE_IPM_samples.*\\.rds$", full.names = TRUE)
  if (!length(cand)) {
    stop(
      "No cached posterior found. Fit the model first (scripts/LEYE_IPM.R or ",
      "scripts/LEYE_IPM_run_parallel.R), or pass -P samples_path:<file>."
    )
  }
  samples_path <- cand[which.max(file.mtime(cand))]
}

# The parallel wrapper's lapply() strips the mcmc.list class, and as.matrix()
# on the resulting plain list silently returns a 3x1 matrix instead of erroring.
# Re-wrap defensively so the report cannot be built on garbage.
samples <- readRDS(samples_path)
if (!inherits(samples, "mcmc.list")) samples <- coda::as.mcmc.list(samples)
mat <- as.matrix(samples)

n_chains <- coda::nchain(samples)
n_draws <- nrow(mat)
stopifnot(n_draws > 100) # guard against the silent-truncation failure above

# Priors on recruitment, as declared in scripts/LEYE_IPM.R. Kept here so the
# prior-vs-posterior comparison below cannot drift from the fitted model.
PRIOR_PI <- c(a = 2.268, b = 13.610)
PRIOR_KAPPA <- c(a = 6.054, b = 6.309)

# omega is a scalar in some fits and a year-indexed vector in others. Resolve
# once here so the rest of the report does not care which it is.
omega_tv <- any(grepl("^omega\\[", colnames(mat)))
omega_draws <- if (omega_tv) mat[, "omega_bar"] else mat[, "omega"]

# tLTRE. Computed here from the SAME posterior loaded above rather than read
# from LEYE_tLTRE_results.rds, so the decomposition cannot silently belong to a
# different fit than the numbers everywhere else in this report. Sourcing with
# SKIP_TLTRE_RUN loads the functions without running the script's own report.
SKIP_TLTRE_RUN <- TRUE
source("scripts/LEYE_tLTRE.R")
tl <- tltre_series(mat, T)
tltre_check(tl) # stops the render if the exact identity ever breaks
tl_level <- tltre_level(tl)
tl_var <- tltre_variance(tl)

driver_labs <- c(
  phi = "Adult survival (φ)",
  omega = "Immigration (ω)",
  f_t = "Productivity, year t",
  f_lag = "Productivity, year t−1",
  rho = "Population structure (ρ)",
  psi1 = "Recruitment at age 1 (ψ1)",
  psi2 = "Recruitment at age 2 (ψ2)"
)
# NB: plain "ψ1" not "ψ₁" -- the Unicode subscript has no glyph in the default
# plotting font and renders as a tofu box on the axis.

q <- function(x) {
  c(
    mean = mean(x),
    med = median(x),
    lwr = quantile(x, .025, names = FALSE),
    upr = quantile(x, .975, names = FALSE)
  )
}
qq <- function(stem, idx) {
  t(vapply(paste0(stem, "[", idx, "]"), function(cn) q(mat[, cn]), numeric(4)))
}

# House table style: one call so every table in the report matches.
tbl <- function(x, title = NULL, subtitle = NULL, dec = 3) {
  g <- gt(x)
  if (!is.null(title)) g <- tab_header(g, title = title, subtitle = subtitle)
  g |>
    fmt_number(columns = where(is.numeric), decimals = dec) |>
    opt_row_striping() |>
    tab_options(
      table.font.size = px(13),
      heading.title.font.size = px(15),
      heading.subtitle.font.size = px(12),
      column_labels.font.weight = "600",
      data_row.padding = px(4),
      table.border.top.style = "none"
    )
}

# Shared plot look.
theme_leye <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey35", size = 11),
      plot.caption = element_text(colour = "grey45", hjust = 0, size = 9),
      axis.title = element_text(colour = "grey25")
    )
}
COL_AD <- "#1f4e79"
COL_REC <- "#2e8b57"
COL_IMM <- "#c0504d"

trans_yr <- ipm_years[-T]
trans_lab <- paste0(trans_yr, "–", trans_yr + 1)
```

::: {.callout-note appearance="simple"}
## How to read this report

Abundance is estimated on a **relative** scale — the counts are a volunteer index,
not a census. Read the *shape* of the trajectory and the growth rate $\lambda$;
the absolute number of birds is not identified and is not reported here.

Recruitment (`psi1`, `psi2`) is a **composite** probability: survive **and**
return to breed locally. It is not juvenile survival, and should not be compared
against published juvenile survival estimates.
:::

# Overview

This integrated population model combines three data sources into one likelihood
so that each informs a shared set of demographic rates.

```{r}
#| label: tbl-sources
#| echo: false

data.frame(
  Source = c("Abundance", "Survival", "Productivity"),
  Data = c(
    "Volunteer wetland counts",
    "Adult mark–resight (CJS)",
    "Nest productivity"
  ),
  Years = c("2014–2025", "2010–2026", "2018–2026"),
  Detail = c(
    sprintf("%d records, %d wetlands, 3 weekly visits", nCount, nWet),
    sprintf("%d individuals, sex + tag-type covariates", cjs$nind),
    sprintf("%d nests, chicks per nest", nProd)
  )
) |>
  tbl("Data sources") |>
  cols_align("left")
```

Recruitment carries a **two-year lag**: chicks are absent from the breeding site
until they recruit, and the 1990s chick data show recruitment split roughly evenly
between ages 1 and 2. Recruits enter the breeding class directly, so there is no
separate juvenile stage — volunteer counts cannot distinguish non-breeding age-1
birds from breeding adults.

$$
R_t = N_{t-1}\,\frac{f_{t-1}}{2}\,\psi_1 \;+\; N_{t-2}\,\frac{f_{t-2}}{2}\,\psi_2
$$

# Headline results

```{r}
#| label: headline
#| echo: false

lam <- rowMeans(mat[, paste0("lambda[", 1:(T - 1), "]")])
geolam <- exp(mean(log(lam)))
geolam_ci <- exp(quantile(log(lam), c(.025, .975)))
Rm <- rowSums(mat[, paste0("N_rec[", 3:T, "]")])
Im <- rowSums(mat[, paste0("Imm[", 3:T, "]")])
share <- Rm / (Rm + Im)

card <- function(value, label, sub, colour) {
  sprintf(
    '<div class="kpi" style="border-top-color:%s"><div class="kpi-v">%s</div>
     <div class="kpi-l">%s</div><div class="kpi-s">%s</div></div>',
    colour, value, label, sub
  )
}
htmltools::HTML(paste0(
  '<div class="kpi-row">',
  card(
    sprintf("%.3f", geolam), "Population growth (λ)",
    sprintf("%.3f – %.3f &nbsp;|&nbsp; %.1f%%/yr", geolam_ci[1], geolam_ci[2], 100 * (geolam - 1)),
    "#c0504d"
  ),
  card(
    sprintf("%.2f", mean(mat[, "phi_ad[1]"])), "Adult survival",
    "annual, untagged female baseline", COL_AD
  ),
  card(
    sprintf("%.3f", mean(mat[, "pi_rec"])), "Local recruitment",
    sprintf("ψ₁ %.3f &nbsp; ψ₂ %.3f", mean(mat[, "psi1"]), mean(mat[, "psi2"])),
    COL_REC
  ),
  card(
    sprintf("%.3f", mean(omega_draws)), "Immigration rate (ω)",
    sprintf("%.0f%% of annual gains", 100 * (1 - mean(share))),
    COL_IMM
  ),
  "</div>"
))
```

The population is declining at roughly
`r sprintf("%.1f%%", abs(100 * (geolam - 1)))` per year. Adult survival is the
best-determined rate in the model. Local recruitment is low, and a substantial
share of annual gains is attributable to immigration — but see
[Recruitment versus immigration] for why that split remains uncertain.

# Population trajectory

```{r}
#| label: fig-trajectory
#| fig-cap: "Estimated breeding population on a relative scale. Ribbon is the 95% credible interval. The absolute level is not identified — read the shape, not the height."

Nt <- as.data.frame(qq("N_tot", 1:T))
Nt$year <- ipm_years

ggplot(Nt, aes(year, med)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = COL_AD, alpha = .15) +
  geom_line(colour = COL_AD, linewidth = .9) +
  geom_point(colour = COL_AD, size = 2.2) +
  scale_x_continuous(breaks = ipm_years) +
  labs(
    title = "Estimated population trajectory",
    subtitle = "Relative scale — interpret shape and growth rate, not absolute abundance",
    x = NULL, y = "Latent abundance (relative)"
  ) +
  theme_leye()
```

```{r}
#| label: fig-lambda
#| fig-cap: "Annual population growth rate. Values below 1 indicate decline. The dashed line marks a stable population."

lamd <- as.data.frame(qq("lambda", 1:(T - 1)))
lamd$transition <- factor(trans_lab, levels = trans_lab)

ggplot(lamd, aes(transition, med)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey45") +
  geom_linerange(aes(ymin = lwr, ymax = upr), colour = COL_AD, linewidth = .7) +
  geom_point(colour = COL_AD, size = 2.2) +
  labs(
    title = "Population growth rate by transition",
    subtitle = sprintf("Geometric mean λ = %.3f (%.3f–%.3f)", geolam, geolam_ci[1], geolam_ci[2]),
    x = NULL, y = expression(lambda)
  ) +
  theme_leye() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Adult survival

`phi_ad` is the baseline rate for **untagged females** — the reference level, and
the rate that drives the population process model, since the population is
overwhelmingly untagged.

```{r}
#| label: fig-survival
#| fig-cap: "Baseline annual adult survival (untagged female). Open points mark transitions with no direct mark–resight data, informed only through the process model."

ad <- as.data.frame(qq("phi_ad", 1:(T - 1)))
ad$transition <- factor(trans_lab, levels = trans_lab)
ad$has_cjs <- trans_yr %in% cjs_int_start

ggplot(ad, aes(transition, med)) +
  geom_linerange(aes(ymin = lwr, ymax = upr), colour = COL_AD, linewidth = .7) +
  geom_point(aes(shape = has_cjs), colour = COL_AD, fill = "white", size = 2.4) +
  scale_shape_manual(values = c(`TRUE` = 21, `FALSE` = 1), guide = "none") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Baseline adult survival",
    subtitle = "Untagged, female reference level",
    x = NULL, y = "Annual apparent survival"
  ) +
  theme_leye() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Covariate effects

Effects are shown as odds ratios against the reference levels (female, untagged).
An odds ratio below 1 means lower survival — or lower resighting probability —
than the reference.

```{r}
#| label: fig-forest
#| fig-cap: "Sex and tag-type effects on survival and resighting. Bars are 95% credible intervals; the dashed line at 1 is no effect."
#| fig-height: 5.2

orq <- function(col, label, panel) {
  x <- exp(mat[, col])
  data.frame(
    term = label, panel = panel,
    med = median(x),
    lwr = quantile(x, .025, names = FALSE),
    upr = quantile(x, .975, names = FALSE),
    pneg = mean(mat[, col] < 0)
  )
}
fo <- rbind(
  orq("bSex_phi[2]", "Male (vs female)", "Adult survival"),
  orq("bSex_phi[3]", "Unknown sex (vs female)", "Adult survival"),
  orq("bTag_phi[2]", "Geolocator (vs untagged)", "Adult survival"),
  orq("bTag_phi[3]", "GPS (vs untagged)", "Adult survival"),
  orq("bTag_p[2]", "Geolocator (vs untagged)", "Resighting"),
  orq("bTag_p[3]", "GPS (vs untagged)", "Resighting")
)
fo$term <- factor(fo$term, levels = rev(unique(fo$term)))

ggplot(fo, aes(med, term)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey45") +
  geom_linerange(aes(xmin = lwr, xmax = upr), colour = COL_AD, linewidth = .7) +
  geom_point(colour = COL_AD, size = 2.4) +
  geom_text(aes(x = upr, label = sprintf("P(<1)=%.2f", pneg)),
    hjust = -0.15, size = 2.9, colour = "grey40"
  ) +
  scale_x_log10(expand = expansion(mult = c(.05, .28))) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(
    title = "Covariate effects (odds-ratio scale)",
    subtitle = "Reference levels: female, untagged",
    x = "Odds ratio (log scale)", y = NULL
  ) +
  theme_leye() +
  theme(panel.grid.major.x = element_line(colour = "grey92"))
```

```{r}
#| label: tbl-groups
#| echo: false

base <- qlogis(mat[, "phi_ad[1]"])
bS <- cbind(0, mat[, "bSex_phi[2]"], mat[, "bSex_phi[3]"])
bT <- cbind(0, mat[, "bTag_phi[2]"], mat[, "bTag_phi[3]"])
grid <- expand.grid(tag = c("None", "Geolocator", "GPS"), sex = c("Female", "Male", "Unknown"))
grid$estimate <- NA_real_
grid$lwr <- NA_real_
grid$upr <- NA_real_
for (i in seq_len(nrow(grid))) {
  v <- plogis(base + bS[, match(grid$sex[i], c("Female", "Male", "Unknown"))] +
    bT[, match(grid$tag[i], c("None", "Geolocator", "GPS"))])
  grid[i, c("estimate", "lwr", "upr")] <- c(mean(v), quantile(v, c(.025, .975)))
}
grid |>
  select(Sex = sex, Tag = tag, Estimate = estimate, Lower = lwr, Upper = upr) |>
  arrange(Sex, Tag) |>
  tbl(
    "Adult survival by sex and tag type",
    "First transition. GPS × unknown-sex has no supporting data."
  ) |>
  cols_align("left", columns = c(Sex, Tag))
```

# Recruitment and immigration

## Local recruitment

`psi1` and `psi2` are the probabilities that a fledged chick survives **and**
returns to breed locally, at age 1 and age 2 respectively. Their priors come from
a separate analysis of 1990s chick cohorts
(`scripts/LEYE_juv90_module.R`) and are informative but free to update.

```{r}
#| label: tbl-recruit
#| echo: false

data.frame(
  Parameter = c("pi_rec (total)", "kappa (fraction at age 1)", "psi1 (age 1)", "psi2 (age 2)"),
  Mean = c(mean(mat[, "pi_rec"]), mean(mat[, "kappa"]), mean(mat[, "psi1"]), mean(mat[, "psi2"])),
  Lower = c(
    quantile(mat[, "pi_rec"], .025), quantile(mat[, "kappa"], .025),
    quantile(mat[, "psi1"], .025), quantile(mat[, "psi2"], .025)
  ),
  Upper = c(
    quantile(mat[, "pi_rec"], .975), quantile(mat[, "kappa"], .975),
    quantile(mat[, "psi1"], .975), quantile(mat[, "psi2"], .975)
  )
) |>
  tbl("Local recruitment", "Composite survival × philopatry — not annual survival") |>
  cols_align("left", columns = Parameter)
```

```{r}
#| label: fig-prior-post
#| fig-cap: "Prior versus posterior for the two recruitment parameters. Movement away from the prior shows the count data carrying information; overlap shows the estimate resting on the 1990s analysis."
#| fig-height: 3.6

grid_x <- seq(0, 1, length.out = 512)
dens <- function(nm, a, b, lab) {
  po <- density(mat[, nm], from = 0, to = 1)
  rbind(
    data.frame(x = grid_x, y = dbeta(grid_x, a, b), which = "Prior", par = lab),
    data.frame(x = po$x, y = po$y, which = "Posterior", par = lab)
  )
}
dd <- rbind(
  dens("pi_rec", PRIOR_PI["a"], PRIOR_PI["b"], "pi_rec (total recruitment)"),
  dens("kappa", PRIOR_KAPPA["a"], PRIOR_KAPPA["b"], "kappa (fraction at age 1)")
)

ggplot(dd, aes(x, y, colour = which, fill = which)) +
  geom_area(position = "identity", alpha = .18, colour = NA) +
  geom_line(linewidth = .8) +
  facet_wrap(~par, scales = "free") +
  scale_colour_manual(values = c(Prior = "grey55", Posterior = COL_REC), name = NULL) +
  scale_fill_manual(values = c(Prior = "grey55", Posterior = COL_REC), name = NULL) +
  labs(title = "Prior versus posterior", x = NULL, y = "Density") +
  theme_leye() +
  theme(legend.position = "top", axis.text.y = element_blank())
```

::: {.callout-important appearance="simple"}
The counts inform the **level** of recruitment but not its **timing**. `pi_rec`
moves appreciably from its prior; `kappa` barely moves. The age-1 versus age-2
split therefore rests entirely on the 1990s chick histories, not on the count
data — and those cohorts are partly simulated (see [Caveats]).
:::

## Recruitment versus immigration {#recruitment-versus-immigration}

```{r}
#| label: fig-gains
#| fig-cap: "Annual gains to the breeding population, split into locally recruited birds and immigrants. Bar heights are posterior means."

gains <- rbind(
  data.frame(
    year = ipm_years[3:T], source = "Local recruits",
    value = colMeans(mat[, paste0("N_rec[", 3:T, "]")])
  ),
  data.frame(
    year = ipm_years[3:T], source = "Immigrants",
    value = colMeans(mat[, paste0("Imm[", 3:T, "]")])
  )
)

ggplot(gains, aes(year, value, fill = source)) +
  geom_col(position = "stack", width = .7) +
  scale_fill_manual(values = c("Local recruits" = COL_REC, "Immigrants" = COL_IMM), name = NULL) +
  scale_x_continuous(breaks = ipm_years[3:T]) +
  labs(
    title = "Sources of annual gains",
    subtitle = sprintf(
      "Local recruitment accounts for %.0f%% of gains (95%% CrI %.0f–%.0f%%)",
      100 * mean(share), 100 * quantile(share, .025), 100 * quantile(share, .975)
    ),
    x = NULL, y = "Birds added"
  ) +
  theme_leye()
```

::: {.callout-warning appearance="simple"}
**This split is genuinely uncertain.** Recruitment and immigration both add birds,
and a relative-scale count index cannot fully separate them. The credible interval
on the recruitment share spans
`r sprintf("%.0f%%", 100 * quantile(share, .025))` to
`r sprintf("%.0f%%", 100 * quantile(share, .975))`. Informative priors on
recruitment narrow this but do not resolve it. Treat the point estimate as
indicative, not as a finding.
:::

# Productivity

```{r}
#| label: fig-fec
#| fig-cap: "Estimated chicks per pair. The final year is estimated from productivity data that has no matching count year and does not feed the population process."

fr <- as.data.frame(qq("f_rate", 1:(T + 1)))
fr$year <- c(ipm_years, max(ipm_years) + 1)
fr$in_model <- c(rep(TRUE, T), FALSE)

ggplot(fr, aes(year, med)) +
  geom_linerange(aes(ymin = lwr, ymax = upr), colour = "grey55", linewidth = .7) +
  geom_point(aes(colour = in_model), size = 2.2) +
  scale_colour_manual(
    values = c(`TRUE` = COL_AD, `FALSE` = "grey60"),
    labels = c(`TRUE` = "In process model", `FALSE` = "Beyond count series"), name = NULL
  ) +
  scale_x_continuous(breaks = fr$year) +
  labs(title = "Productivity", x = NULL, y = "Chicks per pair") +
  theme_leye() +
  theme(legend.position = "top")
```

# What drives population change

A transient LTRE (tLTRE), following Koons et al. (2016, 2017). Two questions that
sound similar and have different answers:

- **Level** — what *sustains* the population? Mean λ splits across survival,
  recruitment and immigration.
- **Variance** — what *drives year-to-year variation* in λ?

Everything below is computed per posterior draw, so each contribution carries a
credible interval rather than being a point estimate at posterior means.

## What sustains the population

```{r}
#| label: fig-tltre-level
#| fig-cap: "Additive contributions to mean λ. Bars are 95% credible intervals; percentages are each pathway's share of λ."
#| fig-height: 2.9

lev <- as.data.frame(tl_level)
lev$pathway <- factor(rownames(tl_level), levels = rev(rownames(tl_level)))
lam_mean <- attr(tl_level, "lambda")[["mean"]]

ggplot(lev, aes(contribution, pathway)) +
  geom_linerange(aes(xmin = lwr, xmax = upr), colour = COL_AD, linewidth = .7) +
  geom_point(colour = COL_AD, size = 2.6) +
  # label at the interval's right edge, not the point, or it lands on the bar
  geom_text(aes(x = upr, label = sprintf("%.1f%%", pct_of_lambda)),
    hjust = -0.25, size = 3.1, colour = "grey35"
  ) +
  scale_x_continuous(expand = expansion(mult = c(.05, .18))) +
  labs(
    title = "Contributions to mean λ",
    subtitle = sprintf("Percentages are share of λ = %.3f", lam_mean),
    x = "Contribution to λ", y = NULL
  ) +
  theme_leye()
```

Survival is the dominant pathway. Recruitment and immigration contribute
comparable, smaller amounts — but see the caveat on separating them below.

## What drives year-to-year variation

::: {.callout-important appearance="simple"}
**Most variation in λ is demographic stochasticity, not environmental variation
in vital rates.** With roughly 25 birds, Poisson noise in who happens to survive,
recruit or arrive is a first-order term. A conventional LTRE reports only the
driver contributions and would present a fraction of the variance as though it
were the whole story.
:::

```{r}
#| label: tbl-tltre-partition
#| echo: false

pp <- as.data.frame(tl_var$partition_pct)
pp$Component <- c(
  "Driver-explained", "Demographic stochasticity", "Cross term"
)
pp[c("Component", "mean", "lwr", "upr")] |>
  setNames(c("Component", "Share of Var(λ) %", "Lower", "Upper")) |>
  tbl(
    "Partition of Var(λ)",
    "The three sum to 100%. Individual shares can exceed it because the cross term is negative.",
    dec = 1
  ) |>
  cols_align("left", columns = Component)
```

The cross term is non-zero because population structure (`ρ`) depends on realized
abundance and so carries the same Poisson noise — the smooth and stochastic parts
of λ are not independent.

```{r}
#| label: fig-tltre-drivers
#| fig-cap: "tLTRE contributions to the driver-explained portion of Var(λ). Bars are 95% credible intervals. Parameters with no year index contribute exactly zero by construction."
#| fig-height: 3.4

ct <- as.data.frame(tl_var$contributions)
ct$driver <- rownames(tl_var$contributions)
ct$label <- driver_labs[ct$driver]
ct$structural_zero <- ct$mean == 0 & ct$lwr == 0 & ct$upr == 0
ct$label <- factor(ct$label, levels = rev(ct$label))

ggplot(ct, aes(mean, label)) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_linerange(aes(xmin = lwr, xmax = upr, colour = structural_zero), linewidth = .7) +
  geom_point(aes(colour = structural_zero), size = 2.6) +
  scale_colour_manual(
    values = c(`FALSE` = COL_AD, `TRUE` = "grey65"), guide = "none"
  ) +
  # phi's upper bound sits past the default break, so widen or it gets clipped
  scale_x_continuous(expand = expansion(mult = c(.04, .08))) +
  labs(
    title = "Contributions to driver-explained variance",
    subtitle = sprintf(
      "Drivers explain %.0f%% of Var(λ); the rest is demographic noise",
      tl_var$partition_pct["driver_explained", "mean"]
    ),
    x = "Contribution to Var(λ)", y = NULL
  ) +
  theme_leye() +
  theme(panel.grid.major.x = element_line(colour = "grey93"))
```

```{r}
#| label: tltre-zeros
#| echo: false
#| results: asis

zeros <- rownames(tl_var$contributions)[
  tl_var$contributions[, "mean"] == 0 & tl_var$contributions[, "upr"] == 0
]
if (length(zeros)) {
  cat(
    "::: {.callout-note appearance='simple'}\n",
    "**", paste(driver_labs[zeros], collapse = ", "), "** contribute exactly ",
    "zero — they are single parameters with no year index, so they cannot vary ",
    "between years *by construction, not by evidence*. A time-varying ",
    "immigration rate was fitted and rejected: it absorbed Poisson noise rather ",
    "than detecting signal (see the README). Separating recruitment from ",
    "immigration needs data that distinguishes chick-origin returns from newly ",
    "arrived unbanded adults.\n:::\n",
    sep = ""
  )
}
```

# Convergence

```{r}
#| label: tbl-rhat
#| echo: false

keyp <- intersect(
  c(
    "pi_rec", "kappa", "psi1", "psi2", "omega", "omega_bar", "sd_omega",
    "phi_pre", "phi_post",
    "mu_f", "lp_base", "logB0",
    "bSex_phi[2]", "bSex_phi[3]", "bTag_phi[2]", "bTag_phi[3]",
    "bTag_p[2]", "bTag_p[3]"
  ),
  colnames(mat)
)
gd <- gelman.diag(samples[, keyp], multivariate = FALSE)$psrf
rh <- data.frame(Parameter = rownames(gd), Rhat = gd[, 1], Upper = gd[, 2])

tbl(rh, "Convergence diagnostics", sprintf("%d chains, %s pooled draws", n_chains, format(n_draws, big.mark = ","))) |>
  cols_align("left", columns = Parameter) |>
  data_color(
    columns = Rhat,
    fn = function(x) ifelse(x > 1.05, "#f8d7da", "#d4edda")
  ) |>
  tab_source_note("Values below 1.05 indicate adequate mixing.")
```

```{r}
#| label: logb0-note
#| echo: false
#| results: asis

if (any(gd[rownames(gd) == "logB0", 1] > 1.05)) {
  cat(
    "::: {.callout-note appearance='simple'}\n",
    "`logB0` is expected to converge poorly. It is the count-model intercept, ",
    "which is non-identifiable against the absolute level of the population — ",
    "only their sum enters the linear predictor. This is structural and does ",
    "not affect $\\lambda$ or any rate reported above.\n:::\n",
    sep = ""
  )
}
```

# Caveats {#caveats}

::: {.callout-important}
## Read before citing any number above

**The 1990s chick cohorts are partly simulated.** Recruitment priors derive from
`scripts/LEYE_juv90_module.R`, whose cohort denominators are real for 1996 only.
For 1995, 1997 and 1998 the number of broods is a random draw with an explicitly
arbitrary mean. Recruitment probability scales inversely with those denominators.
The committed encounter history predates the addition of a random seed and is not
reproducible from the script — treat it as the data of record and regenerate only
deliberately.

**Recruitment and immigration are confounded**, as described above.

**Abundance is relative.** Do not report absolute population size.

**`psi` is not survival.** It is survival combined with returning to breed
locally. Birds that survive but settle elsewhere are excluded; immigration is
modelled separately through `omega`.

**Counts end in 2025** while survival and productivity run to 2026. The extra
survival interval and productivity year are estimated in dedicated pooled
parameters rather than being discarded, and do not drive the trajectory.
:::

# Reproducibility

```{r}
#| label: repro
#| echo: false

data.frame(
  Item = c("Posterior file", "Chains", "Pooled draws", "Model script", "Report rendered"),
  Value = c(
    basename(samples_path),
    as.character(n_chains),
    format(n_draws, big.mark = ","),
    "scripts/LEYE_IPM.R",
    format(Sys.time(), "%Y-%m-%d %H:%M")
  )
) |>
  tbl("Run details") |>
  cols_align("left")
```

To reproduce: fit the model with `scripts/LEYE_IPM_run_parallel.R` (or
`scripts/LEYE_IPM.R` for a serial run), then render this report with

```bash
quarto render reports/leye_ipm_report.qmd
```

<details>
<summary>Session information</summary>

```{r}
#| label: sessioninfo
#| echo: false
sessionInfo()
```

</details>
